---
title: "CCLE Population Example"
output:
  md_document:
    variant: markdown_github
---

##### Prerequisites

The first step is to download the necessary files from the [Cancer Cell Line Encyclopedia](http://www.broadinstitute.org/ccle/home).  For the purpose of the HitWalker2 instance we will use the following files:

```
cell_line_annotations.txt
CCLE_hybrid_capture1650_hg19_NoCommonSNPs_NoNeutralVariants_CDS_2012.05.07.maf.gz
CCLE_NP24.2009_Drug_data_2012.02.20.csv	
CCLE_Expression.Arrays.sif_2012-10-18.txt
CCLE_Expression.Arrays_2013-03-18.tar.gz
```

```{r}

library(hwhelper)
library(RNeo4j)
library(hgu133plus2.db)

```

Note that the `RNeo4j` package needs to be installed from github, which can be done as follows:


```{r eval=FALSE}
library(devtools)
install_github("nicolewhite/RNeo4j")
```

##### Subject Data

First we will make the patient/cellLine to sample mappings.  This involves creation of a subject `data.frame`.  This object should contain the subject name(s), aliases and any
other metadata which will be displayed to the end user.  Users should exercise restraint when specifying metadata as only up to 20 total levels can be displayed at a given time.  In the
example below we will use gender as well as the 'primary site' as our metadata.

```{r}
    
cell.annots <- read.delim("cell_line_annotations.txt", sep="\t", header=T, stringsAsFactors=F)

cell.annots$alias <- ifelse(cell.annots$Cell.line.aliase == "", cell.annots$Cell.line.primary.name,
    paste(cell.annots$Cell.line.primary.name, cell.annots$Cell.line.aliases, sep=" | "))
any(grepl("&", cell.annots))
cell.annots$alias <- gsub("\\s+\\|\\s+", "&", cell.annots$alias)

```

Here we created an 'alias' column which will allow the end user's to search on the subject name as well as the aliases.  The aliases should be delimited by '\&' which is the default delimiter
understood by `load.neo4j`--the internal function that loads these data into Neo4j.

Similarly we can keep/add additional columns which represent the metadata for these subjects.  It is at this point that these data should be recoded and made user-digestable.  Spaces should be
replaced with '_'.  After the subject `data.frame` is ready, we can create a `Subject` object.  By specifying `type` here, it will be used as part of the initial subject/sample display.

```{r}

subject.dta <- cell.annots[,c("CCLE.name", "alias", "Gender", "Site.Primary")]
names(subject.dta) <- c("cellLine", "alias", "gender", "primary_site")

subject.dta$gender[subject.dta$gender == ""] <- "U"
gend.tab <- c(U="Unknown", M="Male", F="Female")
subject.dta$gender <- as.character(gend.tab[subject.dta$gender])

subj <- Subject(subject.info=subject.dta, type="primary_site")

```

##### Variant Data

Variant data can come in several forms.  For cancer next generation DNA sequencing experiments such as this one, the mutation annotation format (MAF) is a common way to summarize the variant
data.  Currently we provide parsing support for this CCLE MAF file though as MAF files are roughly standardized, this should be applicable to other variants as well.

Note the use of the `addSamples` method which is used to automatically add *Subject -> Sample* mapping for this object.

```{r}

maf.file <- "CCLE_hybrid_capture1650_hg19_NoCommonSNPs_NoNeutralVariants_CDS_2012.05.07.maf.gz"

maf.obj <- readMAF.ccle(maf.file)

#add in these sample annotations to the Subject object

all(sampleNames(maf.obj) %in% cell.annots$CCLE.name)#TRUE

#ok, so add them in directly as the cellLine name

addSamples(subj, type="DNASeq") <- maf.obj

```

##### Drug Data

Next we will prepare the drug assay data.  For this purpose we can create a `DrugMatrix` object which consists of a drug x sample matrix as well as a mapping
between drug and gene.  The dowloaded file contains this information, though we need to perform semi-automatic gene symbol curation using the Entrez ID information already contained in the base Neo4j database.

```{r}

library(RNeo4j)

graph = startGraph("http://localhost:7474/db/data/")

drug.dta <- read.csv("CCLE_NP24.2009_Drug_data_2012.02.20.csv", stringsAsFactors=F)

drug.names <- sub("c-", "", unique(drug.dta$Target))

targ.matches <- lapply(drug.names, function(x){

    cypher(graph, "MATCH (gene:EntrezID)-[r:REFERRED_TO]-(symb) WITH gene,symb, r,
                    CASE r.synonyms WHEN null THEN [symb.name] ELSE [symb.name]+r.synonyms END AS gene_syns WHERE ANY(x IN gene_syns WHERE x = {GENE})
                    RETURN gene.name, r.synonyms, symb.name", GENE=x)    
                    
                    })
                    
names(targ.matches) <- drug.names

```

For each of the records, we consider it a match if a single entry was found or if the input symbol matched the display name.  If multiple equally good matches are
found then we will weight them equally.  In addition we will disregard any gene targets not found.

```{r}

gene.ids <- mapply(function(x, name){
  
  if (is.null(x)){
    return(NULL)
  }else if (nrow(x) == 1){
    return(x$gene.name)
  }else if (nrow(x) > 1){
    
    is.match <- x$symb.name == name
    
    if (sum(is.match) == 1){
      
      return(x$gene.name[is.match])
      
    }else if (sum(is.match) > 1){
      return(x$gene.name[is.match])
    }else{
      return(x$gene.name)
    }
    
  }

}, targ.matches, names(targ.matches))

```

The missing targets may refer to gene families more than specific genes and that case would have to be treated specially but we will disregard them for now.

```{r}
print (names(gene.ids[sapply(gene.ids, is.null)]))
```

The automatic approach above is not perfect, for example, the *ABL* gene is likely actually mapped to ABL1 (gene 25) as opposed to *MTTP*.

```{r}
print (gene.ids[sapply(gene.ids, length) > 1])

gene.ids[["ABL"]] <- "25" 

```

The *GS* gene looks ambigous so we will leave it assigned to the two genes reported by NCBI.  Now lets load in the drug data.

```{r}
gene.stack <- stack(gene.ids)
gene.stack$ind <- as.character(gene.stack$ind)

#load in the drug/gene annotation

names(gene.stack) <- c("entrezID", "Target")
gene.stack$weight <- ifelse(gene.stack$Target == "GS", .5, 1)

drug.genes <- drug.dta[,c("Compound", "Target")]

drug.genes <- drug.genes[!duplicated(drug.genes),]

drug.genes.merge <- merge(drug.genes, gene.stack, by="Target", all=T, incomparables=NA, sort=F)

drug.genes.merge <- drug.genes.merge[complete.cases(drug.genes.merge),]
drug.genes.merge <- drug.genes.merge[,c("Compound", "entrezID", "weight")]
names(drug.genes.merge) <- c("drug", "gene", "weight")

#load both the annotations as well as the matrix data

drug.mat <- acast(formula=Compound~CCLE.Cell.Line.Name,data=drug.dta, value.var="IC50..uM.")

drug <- DrugMatrix(mat=drug.mat, mapping=drug.genes.merge)

```

Note the use of the 'weight' column to represent ambiguity in the 'drug' -> 'gene' assignments.  The 'GeneScore' algorithm used internally by HitWalker2 is
based off of the algorithm of (CITATION) and these weights are used internally as part of the summations.

We can also check the implicit *subject -> drug* sample mapping

```{r}

all(sampleNames(drug) %in% cell.annots$CCLE.name)
setdiff(sampleNames(drug), cell.annots$CCLE.name)

```

The `addSamples` method will deal with the discordant samples by creating sample nodes but not the relationships between them.  Which means that:

```{r}
addSamples(subj, type="Drug_Assay") <- drug
```

Implies this:

```{r eval=FALSE}
addSamples(subj) <- data.frame(cellLine=intersect(sampleNames(drug), subject.sample.names),
                                sample=intersect(sampleNames(drug), subject.sample.names),
                                type="Drug_Assay", stringsAsFactors=F)
```

##### Processing expression data

To process the expression data we will choose the top 10% most variable genes (in terms of IQR) and perform a batch correction based on the
Batch information in the array sample information file.  The resulting matrix is then converted back to an ExpressionSet for loading into the
database.

```{r}

sif.dta <- read.delim("CCLE_Expression.Arrays.sif_2012-10-18.txt", sep="\t", header=T, stringsAsFactors=F)

```

```{r eval=FALSE}

library(affy)
library(genefilter)
library(sva)

all.genes <- ReadAffy(path="CCLE_Expression.Arrays_2013-03-18", filenames=list.files(pattern="CEL"))

all(sif.dta$ID %in% sub(".CEL", "", colnames(all.genes)))

all.exprs <- rma(all.genes[,paste0(sif.dta$ID, ".CEL")])

filt.exprs.list <- nsFilter(all.exprs, require.entrez=TRUE,
         remove.dupEntrez=FALSE, var.func=IQR,
         var.cutoff=0.90, var.filter=TRUE,
         filterByQuantile=TRUE, feature.exclude="^AFFX")

filt.exprs <- filt.exprs.list$eset

#Do combat correction based on batch

batch <- sif.dta$Batch
names(batch) <- paste0(sif.dta$ID, ".CEL")

batch <- batch[colnames(filt.exprs)]

int.mod <- model.matrix(~1, data=sif.dta)

comb.exprs <- ComBat(dat=exprs(filt.exprs), batch=batch, mod=int.mod, numCovs=NULL, par.prior=T)

ccle.eset <- ExpressionSet(assayData=comb.exprs, annotation=annotation(filt.exprs))

save(ccle.eset, file="ccle_eset.RData")

```

```{r echo=FALSE}

if (file.exists("ccle_eset.RData")){
  load("ccle_eset.RData")
}
```

For expression, *probeset->gene* mapping info doesn't need to be supplied by an `ExpressionSet` as it can be populated using Bioconductor's annotation mechanism.
As this object doesn't have an annotation package specified, we will manually add it.

As before we still need to supply the mapping between the CEL file names and the subject names and we will do this base on the SIF file.  Instead of supplying `addSamples` an object, we will supply it a `data.frame` with the mapping info as below.

```{r}

sif.dta$ID <- paste0(sif.dta$ID, ".CEL")

all(sif.dta$ID %in% sampleNames(ccle.eset))
all(sif.dta$CCLE.name %in% subj@subject.info$cellLine)

expr.annot <- sif.dta[,c("CCLE.name", "ID")]
names(expr.annot) <- c("cellLine","sample")
expr.annot$type <- "Affy_Expression"

addSamples(subj) <- expr.annot

```

##### Creating the database

Finally we will generate a `HW2Config` object.  To do this we first specify the subject data and the type of gene models (currently only 'entrez').  Next we tell HitWalker2 which
of the datatypes are going to be prioritized ('target') and which datatype(s) is going to be used to provide the basis of the prioritization ('seeds').  Finally we provide the data objects
along with supplied names.

We can then populate the database and create the configuration files.

```{r}

hw2.conf <- makeHW2Config(subject=subj, gene.model="entrez",
        data.types=list(seeds="GeneScore", target="Variants"),
        Expression=HW2exprSet(ccle.eset), GeneScore=drug, Variants=maf.obj)

save(hw2.conf, file="ccle_conf.RData")

populate(hw2.conf)

#configure(hw2.conf)

```

Finally we provide utilities to gain information on the Neo4j graph structure that was created

```{r}

ccle.graph <- compute.graph.structure()

plot(ccle.graph)

```

##### Additional Customization

TODO

```{r}
sessionInfo()
```